### 2012-10-18 doc/prelim/document/chapters/parallel_mc/parallel_mc.tex
finished adding Pauls changes

To quantify the number of particles that leak out of the local domain
they define a leakage fraction, $\lambda$, as:
\begin{equation}
  \lambda =
  \frac{average\ \#\ of\ particles\ leaving\ local\ domain}{total\ 
    of\ \#\ of\ particles\ starting\ in\ local\ domain}\:.
  \label{eq:leakage_fraction}
\end{equation}
For their studies, they assumed that the value of $\lambda$ was
dependent on the total cross section of the system and the same for
all processes for simplicity and load balancing. Using this
relationship and their simple geometry, they were able to relate the
amount of compute time required to transport all source particles in
relation to the number of source particles, the leakage fraction, the
latency and bandwidth of communication, the ratio of local to global
geometry size, and the total cross section of the system.

What they found was that there was indeed an optimum ratio of local
domain to global domain size that results in a minimum total compute
time. They also found that this ratio was only dependent on the
latency and bandwidth constraints of the parallel machine. We expect
this because nearest neighbor communication patterns scale with the
volume to surface sizes of the local domains. The larger the volume,
the more compute time is spent within the local volume and therefore
at some point you are limited by the latency of the system, or the
time it takes to receive the data to operate on. However, at a certain
point, when the surface of those volumes becomes large enough, the
amount of data that must be communicated grows large enough that it
may overcome the bandwidth limitations of the system, resulting in a
decrease in performance. In their numerical experiments, they used the
above geometry and leakage rates based on that computed for a reactor
system for a single material problem with a total cross section
equivalent to that of homogonized light water reactor. Given the
optimal local domain size, they observed for load balanced situations
that nearest neighbor communication costs during transport are a small
fraction of total compute time, leading to the conclusing that domain
decomposition is feasible for large-scale Monte Carlo calculations.

@@

Given a decomposed domain, per Wagner's work
\citep{wagner_hybrid_2010} is clear that domain overlap significantly
reduces the amount of communication required between adjacent domains
as histories move to states that are not owned by the local
processor. How much domain overlap is suitable for matrix inversion
problems? Are we memory limited by large problems such that only so
much overlap is feasible? Are there metrics related to the properties
of the matrix such that we can provide guidelines for selecting the
amount of overlap required? Furthermore, as Wagner and colleagues
observed \citep{wagner_hybrid_2010}, with marginal domain overlap for
reactor problems, the percentage of histories leaving the local domain
can easily be reduced to less than 1\%. Evans and colleagues in their
initial MCSA development typically used only 50 particle histories in
order to compute a correction in an MCSA iteration
\citep{evans_monte_2012}. Per the Central Limit Theorem, that
correlates to a statistical uncertainty of 14.1\% in the correction,
yet as compared to conventional Krylov solvers, MCSA achieved
identical numerical results. If good convergence and numerically
accurate solutions can be achieved with such a large uncertainty in
the correction computation, then perhaps with enough domain overlap
the minimal amount of histories that do transition to non-local states
can be ignored and thus eliminate all communication in the parallel
Monte Carlo adjoint solver transport sequence and create an
embarrassingly parallel method. Given Gropp's work on parallel Krylov
methods that we discussed in
\S~\ref{subsec:projection_method_performance} and the empirical
results of Siegel \citep{siegel_analysis_2012}, we know that these
nearest neighbor computations between adjacent domains do not scale as
well as global reduction operations and therefore we are improving
scaling by eliminating them from the transport sequence. It will be
important to determine if the expectation value bias in the MCSA
correction generated by this approximation will remain within the
bounds of the already high statistical uncertainty for unbiased
estimates, thus providing numerically equivalent results.

>>>>

To quantify the number of particles that leak out of the local domain
they define a leakage fraction, $\lambda$, as:
\begin{equation}
  \lambda =
  \frac{average\ \#\ of\ particles\ leaving\ local\ domain}{total\ 
    of\ \#\ of\ particles\ starting\ in\ local\ domain}\:.
  \label{eq:leakage_fraction}
\end{equation}
For their studies, they assumed that the value of $\lambda$ was
dependent on the total cross section of the system and the same for
all processes for simplicity and load balancing. Using this
relationship and their simple geometry, they were able to relate the
amount of compute time required to transport all source particles to
the number of source particles, the leakage fraction, the latency and
bandwidth of communication, the ratio of local to global geometry
size, and the total cross section of the system.

What they found was that there was indeed an optimum ratio of local
domain to global domain size that results in a minimum total compute
time. They also found that this ratio was only dependent on the
latency and bandwidth constraints of the parallel machine. We expect
this because nearest neighbor communication patterns scale with the
volume and surface sizes of the local domains. The larger the volume,
the more compute time is spent within the local volume and therefore
at some point you are limited by the latency of the system, or the
time it takes to receive the data to operate on. However, at a certain
point, when the surface of those volumes becomes large enough, the
amount of data that must be communicated grows large enough that it
may overcome the bandwidth limitations of the system, resulting in a
decrease in performance. In their numerical experiments, they used the
above geometry and leakage rates based on that computed for a reactor
system for a single material problem with a total cross section
equivalent to that of homogonized light water reactor. Given the
optimal local domain size, they observed for load balanced situations
that nearest neighbor communication costs during transport are a small
fraction of total compute time, leading to the conclusing that domain
decomposition is feasible for large-scale Monte Carlo calculations.

@@

Given a decomposed domain, per Wagner's work
\citep{wagner_hybrid_2010} is clear that domain overlap significantly
reduces the amount of communication required between adjacent domains
as histories move to states that are not owned by the local
processor. How much domain overlap is suitable for matrix inversion
problems? Are we memory limited by large problems such that only so
much overlap is feasible? Are there metrics related to the properties
of the matrix including the eigenvalues and sparsity pattern such that
we can provide guidelines for selecting the amount of overlap
required? Furthermore, as Wagner and colleagues observed
\citep{wagner_hybrid_2010}, with marginal domain overlap for reactor
problems, the percentage of histories leaving the local domain can
easily be reduced to less than 1\%. Evans and colleagues in their
initial MCSA development typically used only 50 particle histories in
order to compute a correction in an MCSA iteration
\citep{evans_monte_2012}. Per the Central Limit Theorem, that
correlates to a statistical uncertainty of 14.1\% in the correction,
yet as compared to conventional Krylov solvers, MCSA achieved
identical numerical results. If good convergence and numerically
accurate solutions can be achieved with such a large uncertainty in
the correction computation, then perhaps with enough domain overlap
the minimal amount of histories that do transition to non-local states
can be ignored and thus eliminate all communication in the parallel
Monte Carlo adjoint solver transport sequence and create an
embarrassingly parallel method. Given Gropp's work on parallel Krylov
methods that we discussed in
\S~\ref{subsec:projection_method_performance} and the empirical
results of Siegel \citep{siegel_analysis_2012}, we know that these
nearest neighbor computations between adjacent domains do not scale as
well as global reduction operations and therefore we are improving
scaling by eliminating them from the transport sequence. It will be
important to determine if the expectation value bias in the MCSA
correction generated by this approximation will remain within the
bounds of the already high statistical uncertainty for unbiased
estimates, thus providing numerically equivalent results.

### 2012-10-18 doc/prelim/document/chapters/research_proposal/research_proposal.tex
incrporating more of Pauls changes. still a few more to go
finished adding Pauls changes

\chapter{Research Proposal}
\label{ch:research_proposal}
To develop the parallel MCSA and FANM methods, a research plan is
proposed to drive their development and quantify their merits and
feasibility as compared to more conventional methods for linear and
nonlinear problems. We seek to rigorously verify both of these methods
and there implementation and will describe the means by which that
will be accomplished. Furthermore, the numerical experiments performed
will be constructed within a software framework that will permit an
agile development strategy. In this chapter we describe that
experimental framework and the progress achieved to date by its
use. In addition, we will a formulate a plan for verifying the new
Monte Carlo methods and prepare a set of numerical experiments that
will determine some of their properties. Finally, we select a
large-scale challenge problem designed to potentially demonstrate the
effectiveness of our new methods in the context of real-world,
production level applications for nuclear reactor analysis.

@@

\section{Monte Carlo Methods Verification}
\label{sec:method_verification}
As we perform our numerical experiments, the results from the new
tools that we are using must be verified against benchmark
results. These benchmark tests should allow use to verify that the new
stochastic solvers that we develop generate the expected numerical
results in parallel for linear and nonlinear problems that have a
known, well-documented solution. In addition, as we explore parallel
strategies for the Monte Carlo methods and introduce bias or
non-reproducible results into our solution estimators, these
verification problems will allow us to determine if that cost is
acceptable. To verify the linear solvers, we will choose the above
2-dimensional heat equation problem in Eq~(\ref{eq:heat_equation}) on
a rectilinear grid for which analytical solutions are available. As a
complement to comparison with the analytical solutions, we may also
check those results against production implementations of Krylov
solvers. For the nonlinear solvers, we employ a new set of equations
with readily available benchmark solutions.

@@

\section{Proposed Numerical Experiments}
\label{sec:numerical_experiments}
Using the verification problems for the linear and nonlinear solvers
as a testing environment, we propose a set of experiments that will be
used to characterize several qualities of parallel Monte Carlo
solvers. At the end of Chapter~\ref{ch:stochastic_methods}, we posed
several research questions related to the parallelization of the
adjoint Neumann-Ulam method and the MCSA method while in
Chapter~\ref{ch:nonlinear_problem} we posed questions related to the
the FANM method. We will generate numerical experiments from these
questions by dividing them into 4 categories: domain overlap studies
for the parallel Neumann-Ulam method, domain replication studies for
the parallel Neumann-Ulam method, parallel performance and numerical
accuracy studies for the MCSA method, and parallel performance and
numerical accuracy studies for the FANM method.

\subsection{Domain Overlap Studies for the Parallel Neumann-Ulam
  Method}
\label{subsec:domain_overlap_studies}
Characterizing and understanding the domain overlap behavior of a
parallel Neumann-Ulam method will be critical to efficient
implementations of other solvers that will be built on top of these
methods. In the reactor physics work that we reviewed, the amount
domain overlap used could be directly correlated to physical
parameters of the system, namely the dimension of repeated geometric
structures and the mean free path of neutrons in the system. We
therefore seek analogous measurements for the general linear
system. These measurements will characterize how far from its starting
state a random walk sequence moves as a function of properties of the
linear system. Our experimental framework will allow us to fix various
properties of the system while varying others including the size of
the linear system, its spectral radius, and other parameters related
to the Eigenvalues and conditioning of the linear operator. By
varying the domain overlap and measuring parallel performance with
respect to these properties, we can develop guidelines for setting the
overlap with respect to measurable properties of the linear
operator. In addition, varying the amount of overlap will allow us to
measure the additional memory costs incurred.

>>>>

\chapter{Research Proposal}
\label{ch:research_proposal}
To develop the parallel MCSA and FANM methods, a research plan is
proposed to drive their development and quantify their merits and
feasibility as compared to more conventional methods for linear and
nonlinear problems. We seek to rigorously verify both of these methods
and their implementation and will describe the means by which that
will be accomplished. Furthermore, the numerical experiments performed
will be constructed within a software framework that will permit an
agile development strategy. In this chapter we describe that
experimental framework and the progress achieved to date by its
use. In addition, we will a formulate a plan for verifying the new
Monte Carlo methods and prepare a set of numerical experiments that
will determine some of their properties. Finally, we select a
large-scale challenge problem designed to potentially demonstrate the
effectiveness of our new methods in the context of real-world,
production level applications for nuclear reactor analysis.

@@

\section{Monte Carlo Methods Verification}
\label{sec:method_verification}
As we perform our numerical experiments, the results from the new
tools that we are using must be verified against benchmark
results. These benchmark tests should allow use to verify that the new
stochastic solvers that we develop generate the expected numerical
results in parallel for linear and nonlinear problems that have a
known, well-documented solution. In addition, as we explore parallel
strategies for the Monte Carlo methods and introduce bias or
non-reproducible results into our solution estimators, these
verification problems will allow us to determine if that cost is
acceptable. To verify the linear solvers, we will choose the above
2-dimensional heat equation problem in Eq~(\ref{eq:heat_equation}) on
a rectilinear grid for which analytical solutions are available. As a
complement to comparison with the analytical solutions, we may also
check those results against production implementations of Krylov
solvers. For the nonlinear solvers, we employ a new set of equations
with readily available benchmark solutions. In both cases, we
verification will be determined by comparing the locations of global
minima and maxima in the solution as well as comparison of solution
vector norms with respect to a specified tolerance.

@@

\section{Proposed Numerical Experiments}
\label{sec:numerical_experiments}
Using the verification problems for the linear and nonlinear solvers
as a testing environment, we propose a set of experiments that will be
used to characterize several qualities of parallel Monte Carlo
solvers. At the end of Chapter~\ref{ch:stochastic_methods}, we posed
several research questions related to the parallelization of the
adjoint Neumann-Ulam method and the MCSA method while in
Chapter~\ref{ch:nonlinear_problem} we posed questions related to the
FANM method. We will generate numerical experiments from these
questions by dividing them into 4 categories: domain overlap studies
for the parallel Neumann-Ulam method, domain replication studies for
the parallel Neumann-Ulam method, parallel performance and numerical
accuracy studies for the MCSA method, and parallel performance and
numerical accuracy studies for the FANM method.

\subsection{Domain Overlap Studies for the Parallel Neumann-Ulam
  Method}
\label{subsec:domain_overlap_studies}
Characterizing and understanding the domain overlap behavior of a
parallel Neumann-Ulam method will be critical to efficient
implementations of other solvers that will be built on top of these
methods. In the reactor physics work that we reviewed, the amount
domain overlap used could be directly correlated to physical
parameters of the system, namely the dimension of repeated geometric
structures and the mean free path of neutrons in the system. We
therefore seek analogous measurements for the general linear
system. These measurements will characterize how far from its starting
state a random walk sequence moves as a function of properties of the
linear system. Our experimental framework will allow us to fix various
properties of the system while varying others including the size of
the linear system, its spectral radius, its sparsity, its asymmetry,
and other parameters related to the Eigenvalues, conditioning, and
structure of the linear operator. By varying the domain overlap and
measuring parallel performance with respect to these properties, we
can develop guidelines for setting the overlap with respect to
measurable properties of the linear operator. In addition, varying the
amount of overlap will allow us to measure the additional memory costs
incurred.
